Kelp morphometric properties and the link to hydrodynamic modelling

### 2018-09-21 chapter_1.Rmd
more additions and edits to the wave exposure section

Other than temperature, wave exposure is also recognised as an important driver of the marine environment, and macroalgae are not exception. Wave exposure has been shown to play a role in determing distribution, abundance, diversity, composition, growth [@Cousens1982] and productivity [@pedersen2012a] of macroalgae communties. For example, the width, vertical zonation and diversity of algal communities often change predictably along gradients of wave exposure. Wave exposure may also drive macroalgae communties indirectly through the alteration in affect of another environmental driver. For instance, increasing degrees of exposure may positively influence the amount of area available to trap light on macroalgal fronds, as well as increasing nutrient uptake due to increased turbulence in the boundry layer around the frond [@lobban1994]. The most important direct effect of wave exposure on macroalgal communities is through mechanical dislodgment, which ultimately leads to expiration. Wave exposure is a complex abiotic variable which varies spatially and temporarly in the marine enviroment. Furthermore, the degree to which a macroalgae community is exposed which depend on local site characteristics, such as bathymetry and local wind patterns. Despite this fact, macrolagae have been able to persist in often harsh and variable wave environments. Macroalage are sessile organisms and incapabable of migrating when local conditions become unsuitable. Therefore, macroalgae must adapt to the local wave climate in order to persist and survive, and achieve this through morphological variation. This morphological variation allows macroalgae to reduce the amount of drag experienced during times of high wave energy or pulse distrubance events, such as storms. This may be achieved through different strategies and are species-specific, which can be directly attributed the high diversity in morphological characters of algal communities. For example some algae have fronds and others are articulated corraline, and therefore these species would need to adpat their morphology differently in order to persist. @moss1948 investigated the anatomy, chemical composistion of *Fucus spiralis* at three sites that varied in wave exposure (sheltered, medium exposure and exposed). The authors found that individuals in exposed sites showed less branching of thalli as well as variation in organic nitrogen, mannitol, laminarian and alginic acid concentrations.

@@

Traditional ecological measures of wave exposure usually incorportates integrative measures of hyrdodynamic conditions at a particular site. More specifically, it is the integration of mechanical processes and the influence that is has on the ecology of nearshore communities.

Wave exposure may be modelled through various methods which range from simple cartographic to more advanced numerical wave models. Cartographical models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurments for larger areas, and has been suggested as a method for predicting macroalgal community structure [@burrows2008]. An example of such a model is the "BioEx model" which was developed by @baardseth1970 to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open). This method has been used in mapping of marine coastal biodiveristy [@Rinde2004].

>>>>

Other than temperature, wave exposure is also recognised as an important driver of the marine environment, and macroalgae are not exception. Wave exposure has been shown to play a role in determing distribution, abundance, diversity, composition, growth [@Cousens1982] and productivity [@pedersen2012a] of macroalgae communties. For example, the width, vertical zonation and diversity of algal communities often change predictably along gradients of wave exposure. Wave exposure may also drive macroalgae communties indirectly through the alteration in affect of another environmental driver. For instance, increasing degrees of exposure may positively influence the amount of area available to trap light on macroalgal fronds, as well as increasing nutrient uptake due to increased turbulence in the boundry layer around the frond [@lobban1994]. The most important direct effect of wave exposure on macroalgal communities is through mechanical dislodgment, which ultimately leads to expiration. Wave exposure is a complex abiotic variable which varies spatially and temporarly in the marine enviroment. Furthermore, the degree to which a macroalgae community is exposed, is dependent on local site characteristics, such as bathymetry and local wind patterns. Despite this fact, macrolagae have been able to persist in often harsh and variable wave environments. Macroalage are sessile organisms and incapabable of migrating when local conditions become unsuitable. Therefore, macroalgae must adapt to the local wave climate in order to persist and survive, and achieve this through morphological adaptation. The morphology of macroalgae are not fixed gentic traits. A study by @?? showed that *species name* was able to rapidly adapt to the local wave environment when transplanted from a low wave exposure site to a high wave exposure site. Advances in genetic techniques and taxonomy have revealed that species delination based on morphology has been inaccurate, and organisms that were once considered two seperate species are actually one species.**insert example here**. The morphological adaption that macroalgae display are therefore driven by local or site conditions, and therefore individuals must be morphologically flexible to persist in stochastic environments. This may be achieved through different strategies and are species-specific, which can be directly attributed the high diversity in morphological characters of algal communities. For example some algae have fronds and others are articulated corraline, and therefore these species would need to adapt their morphology differently in order to persist. @moss1948 investigated the anatomy, chemical composistion of *Fucus spiralis* at three sites that varied in wave exposure (sheltered, medium exposure and exposed). The authors found that individuals in exposed sites showed less branching of thalli as well as variation in organic nitrogen, mannitol, laminarian and alginic acid concentrations. **insert more examples here**.

@@

Wave exposure may be modelled through various methods which range from simple cartographic to more advanced numerical wave models. Traditional ecological measures of wave exposure usually incorportates integrative measures of hyrdodynamic conditions at a particular site. More specifically, it is the integration of mechanical processes and the influence that is has on the ecology of nearshore communities. Cartographical models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurments for larger areas, and has been suggested as a method for predicting macroalgal community structure [@burrows2008]. An example of such a model is the "BioEx model" which was developed by @baardseth1970 to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open). This method has been used in mapping of marine coastal biodiveristy [@Rinde2004].
